---
title: "NEON download"
author: "Kari Norman"
date: "10/15/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(neonUtilities)
library(piggyback)
```

Download data
```{r}
beetle_dir <- here::here("data", "beetles_raw.rds")
if (file.exists(beetle_dir)){
  load(beetle_dir)
} else{
  beetles_raw <- loadByProduct("DP1.10022.001", check.size = FALSE)
  saveRDS(beetles_raw, beetle_dir)
}
```

Grab necessary columns from different tables. Important column info:
* sampleID: unique ID for each combination of domainID, siteID, plotID, trapID, and collectDate (unique ID for bet_fielddata table)
* subsampleID: unique ID for each group of individuals, multiple per sample (unique ID for bet_sorting); unclear why there are muliple entries for the same classification
* individualID: unique ID for each individual identified by taxonomists (unique ID for bet_parataxonimist table)

There are multiple levels of taxonomic identification, need to decide which level of rigor we care about (NEON id often just to family, taxonomist ID to species but only for a subset, DNA ID for even smaller subset).

```{r}
data <- beetles_raw$bet_fielddata %>%
  filter(sampleCollected == "Y") %>% #there's an entry for every trap, whether or not they got samples, only want ones with samples
  select(sampleID, domainID, siteID, plotID, trapID, collectDate) %>%
  left_join(beetles_raw$bet_sorting %>%
              filter(sampleType %in% c("carabid", "other carabid")) %>% #only want carabid samples, not bycatch
              select(sampleID, subsampleID, sampleType, taxonID, scientificName, 
                     taxonRank, individualCount, identificationQualifier),
            by = "sampleID") %>%
  filter(!is.na(subsampleID)) %>% #even though they were marked a sampled, some collection times don't acutally have any samples
  mutate(year = lubridate::year(collectDate),
         month = lubridate::month(collectDate),
         day = lubridate::day(collectDate))
 
```

Join taxonomic data from pinning with the sorting data
```{r}
# Replace sorting taxon info with pinning taxon info (people that pin specimens are more experienced with taxonomy), where available
data_pin <- data %>%
  left_join(beetles_raw$bet_parataxonomistID %>% 
              select(subsampleID, individualID, taxonID, scientificName, 
                     taxonRank,identificationQualifier), 
            by = "subsampleID") %>%
  mutate_if(is.factor, as.character) %>%
  mutate(taxonID = ifelse(is.na(taxonID.y), taxonID.x, taxonID.y)) %>%
  mutate(taxonRank = ifelse(is.na(taxonRank.y), taxonRank.x, taxonRank.y)) %>%
  mutate(scientificName = ifelse(is.na(scientificName.y), scientificName.x, scientificName.y)) %>%
  mutate(identificationSource = ifelse(is.na(scientificName.y), "sort", "pin")) %>%
  mutate(identificationQualifier = ifelse(is.na(taxonID.y), identificationQualifier.x, identificationQualifier.y)) %>%
  select(-ends_with(".x"), -ends_with(".y"))

#some subsamples weren't fully ID'd by the pinners, so we have to recover the unpinned-individuals
lost_indv <- data_pin %>% 
  filter(!is.na(individualID)) %>%
  group_by(subsampleID, individualCount) %>%
  summarise(n_ided = n_distinct(individualID)) %>% 
  filter(n_ided < individualCount) %>%
  mutate(unidentifiedCount = individualCount - n_ided) %>%
  select(subsampleID, individualCount = unidentifiedCount) %>%
  left_join(select(data, -individualCount), by = "subsampleID") %>%
  mutate(identificationSource = "sort")

#add unpinned-individuals back to the pinned id's, adjust the individual counts so pinned individuals have a count of 1
data_pin <- data_pin %>%
  mutate(individualCount = ifelse(identificationSource == "sort", individualCount, 1)) %>%
  bind_rows(lost_indv)
```

Join expert data to existing pinning and sorting data
```{r}
# Add expert taxonomy info, where available
data_expert <- left_join(data_pin,
  select(
    beetles_raw$bet_expertTaxonomistIDProcessed,
    individualID,
    taxonID,
    scientificName,
    taxonRank,
    identificationQualifier),
  by = 'individualID', na_matches = "never"
) %>% distinct()
 
#For some reason there are ~10 individualID's for which experts ID'd more than one species, that's why 10 additional records are added by the join
#beetles_raw$bet_expertTaxonomistIDProcessed %>% group_by(individualID) %>% filter(n_distinct(taxonID) > 1) %>% View()

# Replacement old taxon info with expert info, where available
data_expert <- data_expert %>%
  mutate_if(is.factor, as.character) %>%
  mutate(taxonID = ifelse(is.na(taxonID.y), taxonID.x, taxonID.y)) %>%
  mutate(taxonRank = ifelse(is.na(taxonRank.y), taxonRank.x, taxonRank.y)) %>%
  mutate(scientificName = ifelse(is.na(scientificName.y), scientificName.x, scientificName.y)) %>%
  mutate(identificationSource = ifelse(is.na(scientificName.y), identificationSource, "expert")) %>%
  mutate (identificationQualifier = ifelse(is.na(taxonID.y), identificationQualifier.x, identificationQualifier.y)) %>%
  select(-ends_with(".x"), -ends_with(".y"))

beetles_data <- as_tibble(data_expert)
```

Get raw counts table
```{r}
beetles_counts <- beetles_data %>%
  select(-c(subsampleID, sampleType, individualID, identificationSource, identificationQualifier)) %>%
  group_by_at(vars(-individualCount)) %>%
  summarise(count = sum(individualCount)) %>%
  ungroup()

saveRDS(beetles_counts, "data/beetles_counts.rds")
```
